Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-14 Manuscript/manuscript.tex

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The parameters are available at \cite{??} \todo{To be added}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. Chloride counter ions were used
in simulations, because parameters for bromide, the counterion in the experimental data \cite{scherer89},
were not available in the standard Gromacs files for Amber force field.
The Lipid14 model was used for POPC.
The first 20~ns of the total simulation time of 200~ns
was considered as an equlibration time and was omitted from the analysis.
A reasonable lipid neighbor exchange occured during the simulation.

@@

\subsection{Response of POPC head groups to bound charge}\label{section:boundCHARGE}

@@

\subsection{Cation binding affinities to POPC}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=7.5cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    Changes of head group order parameters of POPC bilayer as a function of CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from Ref. \cite{catte16}.
  }
  \todo{I think that we must show the NaCl results in the main paper. I think that the best would
    be to use similar two column format, which we had before. This is used also in NMRlipids II and makes the
    comparison easier} \\
  \todo{The DPPC and POPC should be shown and properly labelled as we had previously.} \\
  \todo{The labeling should be consistent with previous figures, i.e., experiments with diamonds.} \\
  \todo{Start x-axis scale from -1 to make the point in zero fully visible.} \\
  \todo{Start y-axis scale of the two top figures from -0.13 to remove the empty space.} \\
  \todo{y-axis scale of the bottom figure from 59 to 78 to remove the empty space.} \\
  \todo{Empty space between figures and from the right column could be reduced.} \\
  \todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    the value of $g_3$ order parameter of DPPC was -0.214 
    in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    at the C-2 segments of both fatty acyl chains.''
  }
\end{figure}

The changes of the lipid bilayer headgroup order parameter from different simulations and
experiments \cite{akutsu81,altenbach84} are shown in Fig.~\ref{fig:delta_ordPar_CaCl}
as a function of NaCl and CaCl$_2$ concentrations.
%Responses to NaCl concentrations are in SI, Fig.~S\ref{fig:delta_ordPar_NaCl}.
These results can be used to compare the ion binding affinities to lipid bilayers
between simulations and experiments by using the electrometer concept, because 
the order parameters decrease proportionally to the amount of bound positive charge~\cite{seelig87,catte16}. 
The recent comparison of different simulation models to the experimental data
revealed that most models significantly overestimate the Na$^+$ ion binding to PC
lipid bilayers and that none of the available models correctly reproduce
the details of the binding of Ca$^{2+}$ ions \cite{catte16}.
A positive exception was the Lipid14 model \cite{dickson14} simulated with \AA{}qvist ions,
as seen from the data replotted from Ref.~\citenum{catte16} in Fig.~\ref{fig:delta_ordPar_CaCl}.
The model reproduced the experimentally measured small order parameter changes with NaCl
and imperceptible Na$^+$ binding to PC bilayers \cite{akutsu81,altenbach84}.
However, the changes of the headgroup order parameters as a function of CaCl$_2$ 
concentration were overestimated by the same combination of force field parameters.

@@

To improve the cation binding behaviour to POPC lipid bilayer we implicitly included the electronic
polarization to the Lipid14 model by using the ECC approach as described in section \ref{section:ecc}.
The ECC-lipid model simulations with the ECC-ion models \cite{??}
exhibit a good agreement with the experimental changes of 
the lipid headgroup order parameters as a function of NaCl and CaCl$_2$ concentrations
in Fig.~\ref{fig:delta_ordPar_CaCl}. Since also the headgroup order parameter
response the to the bound positive charge the ECC-lipid model
was in good agreement with experiments in section \ref{section:boundCHARGE},
we conclude that the model correctly reproduces the binding affinity of
Na$^{+}$ and Ca$^{2+}$ ions to POPC lipid bilayer.

The density distributions of water and ions along the membrane normal
are shown in Fig.~\ref{fig:cacl-dens}.
\todo{Densities and relative surface excess from NaCl systems to be added.
  The discussion below to be finished after this.}
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
To quantify the binding affinities in different models we calculated 
the relative surface excess from Eq.~\ref{surfexcess}. The value for Calcium
in the ECC-lipid model, $\Gamma_i^w = 0.07 \pm 0.01 \rm{nm}^{-2}$, is significantly
smaller than the values for the Lipid14 model with \AA{}qvist ions, $\Gamma_i^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_i^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.

%\todo{Below analysis is done in a stupid way to get some idea. I would fine it useful to
%  do this analysis by using the density profiles, but it is not necessary.
%  The reason why I want to do something like this, is that it would be useful to have
%  some easier way to estimate the correct binding affinity than the electrometer concept,
%  which is quite tedious to apply in practise (simulations with different concenrations,
%  cationic surfactant check, etc.). I wuold like to be able to estimate much faster from a
%  simulation if the binding affinity is reasonable or not. This may not be the correct way for
%that anyway.}
%Rough estimates for the free energy difference between bound and unbound cation
%are given by
%\begin{equation}
%  \Delta G = -{\rm k_b T}\log(\frac{p_i}{p_o}),
%\end{equation}
%where $p_o$ and $p_i$ are estimated from the Ca$^{2+}$densities in bulk water and
%in the maximum density in bilayer, respectively. The density profiles in
%Fig. \ref{fig:cacl-dens} give $\sim 0.8$ k$_{\rm b}$T for the free energy difference
%between bound and unbound Ca$^{2+}$ ions in ECC-lipid model and $\sim 1.4$ k$_{\rm b}$T
%in Lipid14 with \AA{}qvist.

@@

In conclusion, the results suggest that 
the ECC-lipid model together with the ECC-ion parameters gives a realistic
description for the binding affinities of Na$^+$ and Ca$^{2+}$ ions to POPC lipid
bilayer. Furthermore, the overestimated headgroup order parameter changes of POPC in the Lipid14
model as a function of CaCl$_2$ concentration arise from both, the overestimated
binding affinity and sensitivity of the headgroup tilt to the bound positive charge.
This probably applies also to the other lipid models considered in the previous study~\cite{catte16},
emphasizing the importance of the comparison of the lipid headgroup order parameter response
to the bound charge between simulations and experiments as done in section \ref{section:boundCHARGE}.

\subsection{Binding stoichiometry}

@@

This section is rough and will likely require editing. 
Data -- up to noted exceptions -- shall be all there, however. 
\todo{SAMULI: I think the we should make clear that the current binding
  constants in, e.g., Marsh's handbook are based on the headgroup order parameter
  changes intepreted with ternary complex binding model. I.e. the experimental raw data
  is excatly the same as we have in this work. Our simulations enable much more versatile interpretation
  of the binding phenomena. I think that this is one the main points of this work, so this comment
  applies to the previous section, introduction, abstract and conclusions as well.
  JOE: agree, this would be a good statement in general. I'll try to put it into the text.
}

Binding stoichiometry of \ce{Ca^{2+}} and POPC was 
thoroughly studied in the experimental work~\cite{altenbach84},
in which the head group order parameter changes to cation binding are determined.
Several binding models were proposed and tested of which only one,
ternary complex binding model, provided a good fit of the experimental observations.

Simulations allow us to directly evaluate the stoichiometry 
by calculating relative propensities of various \ce{Ca^{2+}}:$n\times$POPC clusters
by evaluating contacts between cations and lipids with a cut off radius $0.3\,\mathrm{nm}$. 
In Figure~\ref{fig:cacl_complexes} we see that ternary complex is indeed 
the most probable binding mode of calcium at 285~mM concentration. 
Apart from this complex, we also find complexes with 1~and 3~lipids
occuring with only a slightly lower but similar probability.
The fractions of \ce{Ca^{2+}}:$n\times$POPC complexes at 285~mM concetration 
are then in order:
$42\%$ for two lipids,
$30\%$ for one lipid, and
$28\%$ for three lipids.

Several binding models were proposed and tested~\cite{altenbach84} of which only one,
ternary complex binding model, provided a good fit of the experimental observations. 
In such a model, it is assumed that \ce{Ca^{2+}} cations bind to a POPC membrane 
with a stoichiometry 2~POPC:1~\ce{Ca^{2+}}.
In a later work~\cite{macdonald87}, 
a Langmuir adsorption model (i.e. stoichiometry 1~POPC:1~\ce{Ca^{2+}}) 
was found to provide as good fit as ternary complex model, 
when only low concentrations of \ce{CaCl2} are considered.
Ternary complex model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
The symmetry of the distribution of complexes from simulation --  
i.e. almost equal probabilities of complexes with 1 or 3 lipids 
that behave in the total average picture as complexes with 2 lipids -- 
provides clues why ternary complex binding model 
fits both simulation and experimental results relatively well, 
although it is apparently incorrect.

@@

In addition, we estimated relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}}. 
The dominant contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
Such a finding is based on the calculated
number of contacts per lipid between \ce{Ca^{2+}} and POPC oxygen atoms, 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O}$,
in which a contact between the two atoms is counted 
if their distance is less than or equal $0.3 \, \mathrm{nm}$ 
(robustness of this definition discussed in reference~\citenum{javanainen17}). 
% Numbers from simulation with 346mM conc. CaCl2
The average number of contacts per lipid between \ce{Ca^{2+}} and any oxygen atom in POPC is 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.134 $,    % = 17.1/128 
whereas if only phosphate oxygen atoms are considered 
the average number of contacts decreases only by a tiny amount to 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.133 $,    % = 17.0/128 . 
and if only carbonyl oxygen atoms are considered 
(phosphate oxygens contribute, but are not counted as contacts),
this value is merely
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.033 $.    % = 4.3/128 . 
% Numbers from simulation with 980/869mM conc. CaCl2
%The average number of contacts per lipid between \ce{Ca^{2+}} and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $,    % = 33.8/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.262 $,    % = 33.5/128 . 
%and if only carbonyl oxygen atoms are considered 
%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.066 $.    % = 8.4/128 . 
This means that calcium is in contact with a phosphate oxygen for 99\% of the time it is bound to the membrane, 
whereas only 25\% of the time accounts for contact with a carbonyl oxygen,
and purely carbonyl bound states account for less than 1\% of configurations. 
We corroborate this finding with the probability isodensity contours in Fig.~\ref{fig:volmaps}, 
which are easier to interpret.

The residence time of \ce{Ca^{2+}} bound to a POPC membrane 
is experimentally estimated to be lower than $10\,\mu\mathrm{s}$~\cite{altenbach84}. 
From a recent theoretical work with long simulations 
(Fig. S17 in \citenum{javanainen17} -- C36 with MK Ca2+, any other model, either different cation or Slipids/C36, just such calcium)
this time can be roughly estimated %-- get the simulation from \url{https://doi.org/10.5281/zenodo.259376} and analyze it -- 
in the order of $1\,\mu\mathrm{s}$~\cite{javanainen17}.

Improve the above statement so that it contains the reason why I estimate it to this value.
Quick result (Matti C36 with CaClMK, 800ns total):
90\% of bound residence time is smaller than 787 ns, 
The longest is 800 ns (the simulation length)
and it is only less than
60\% of bound residence time is smaller than 400 ns -- half of the simulation length 
and it is only less than
50\% of bound residence time is smaller than 290 ns.
Apparently the exchange of calcium between membrane and solvent 
is apparently limited by the length of the simulation.

This is in contrast to our model, ECC-lipids, 
which estimates the residence time of calcium bound to a POPC membrane to be 
smaller than $53\,\mathrm{ns}$ for 90\% of the time a cation is in a membrane-bound state. 
i.e. at least two orders of magnitude lower than previous estimates.
In addition, the longest observed residence time in our simulation is $120\,\mathrm{ns}$, 
which is well below the length of our simulations.
This is unfortunately not the case of many theoretical works,
where the residence time cannot be estimated,
or the simulation has not even decorrelated from it's 
initial condition with a population of bound/unbound cations \cite{catte16}, 
because it is limited by the simulation length, 
even though it can be very long (order of $\mu\mathrm{s}$) in some works, e.g. reference~\citenum{javanainen17}. 
Increasing time scales to the order of $\mu\mathrm{s}$ 
doesn't help to alleviate such a problem in current lipid models \cite{javanainen17}. %and also \cite{catte16}, but same simulation was used. 
Such a finding changes the point of view of calcium binding from
very tight long-term stable binding with rare exchanges to 
a relatively frequent exchange of cations 
in equilibrium between membrane and solvent.

>>>>

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The parameters are available at \cite{??} \todo{To be added}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. 
Chloride counter ions were used
in simulations, because parameters for bromide, the counterion in the experimental data \cite{scherer89},
were not available in the standard Gromacs files for Amber force field.
\todo{JOE: We cannot present an intricate cool new model for a lipid and claim at the same time that we can't simulate bromide. }
The Lipid14 model was used for POPC.
The first 20~ns of the total simulation time of 200~ns
was considered as an equlibration time and was omitted from the analysis.
A reasonable lipid neighbor exchange occured during the simulation.

@@

\subsection{Calibration of lipid electrometer:
            Response of POPC head groups to bound charge}\label{section:boundCHARGE}

@@

\subsection{Cation binding affinities to POPC membrane validated through lipid electrometer}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/PN_angle_OrdPars-A-B-g3_L14-ECCL17_q80_sig89_CaCl.pdf}
  \caption{\label{fig:delta_ordPar_CaCl}
    Changes of head group order parameters of POPC bilayer as a function of CaCl$_2$ concentrations
    are shown from simulations with different force fields together with experimental data 
    (DPPC \cite{akutsu81} and POPC \cite{altenbach84}). 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from Ref. \cite{catte16}.
  }
  \todo{I think that we must show the NaCl results in the main paper. I think that the best would
    be to use similar two column format, which we had before. This is used also in NMRlipids II and makes the
    comparison easier} \\
  \todo{The DPPC and POPC should be shown and properly labelled as we had previously.} \\
  \todo{The labeling should be consistent with previous figures, i.e., experiments with diamonds.} \\
  \todo{Start x-axis scale from -1 to make the point in zero fully visible.} \\
  \todo{Start y-axis scale of the two top figures from -0.13 to remove the empty space.} \\
  \todo{y-axis scale of the bottom figure from 59 to 78 to remove the empty space.} \\
  \todo{Empty space between figures and from the right column could be reduced.} \\
  \todo{Experimental values from \cite{akutsu81} to be put in the $g_3$ figure:
    the value of $g_3$ order parameter of DPPC was -0.214 
    in the absence of ions [average of two closely spaced splittings] and -0.211 in the presence
    of 0.35 M CaCl$_2$ (at 59 Celcius). No effect of ions could be detected on DPPC bilayers labeled
    at the C-2 segments of both fatty acyl chains.''
  }
\end{figure}

\begin{figure}[tbp]
  \centering
  \includegraphics[width=8.0cm]{../Fig/ipython_nb/OrdPars-A-B_L14-ECCL17_q80_sig89_NaCl.pdf}
  \caption{\label{fig:delta_ordPar_NaCl}
    Changes of head group order parameters $\alpha$ and $\beta$ of POPC bilayer as a function of NaCl concentrations
    are shown from simulations with different force fields together with experimental data \cite{akutsu81}. 
    Ion concentrations in bulk water are shown in x-axis. 
    Values from simulations are calculated from the of cation number density $C_{np}$
    from the region at the simulatin box edge with the constant ion concentration as [ion]=$C_{np}/0.602$.
    Simulation data with Lipid14 and \AA{}qvist ion parameters is taken directly from Ref. \cite{catte16}.
  }
\end{figure}

The changes of the lipid bilayer headgroup order parameter from different simulations and
experiments \cite{akutsu81,altenbach84} are shown in Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_NaCl}
as a function of NaCl and CaCl$_2$ concentrations.
%Responses to NaCl concentrations are in SI, Fig.~S\ref{fig:delta_ordPar_NaCl}.
These results can be used to compare the ion binding affinities to lipid bilayers
between simulations and experiments by using the electrometer concept, because 
the order parameters decrease proportionally to the amount of bound positive charge~\cite{seelig87,catte16}. 
The recent comparison of different simulation models to the experimental data
revealed that most models significantly overestimate the Na$^+$ ion binding to PC
lipid bilayers and that none of the available models correctly reproduce
the details of the binding of Ca$^{2+}$ ions \cite{catte16}.
A positive exception was the Lipid14 model \cite{dickson14} simulated with \AA{}qvist ions,
as seen from the data replotted from Ref.~\citenum{catte16} in Fig.~\ref{fig:delta_ordPar_CaCl}.
The model reproduced the experimentally measured small order parameter changes with NaCl
and imperceptible Na$^+$ binding to PC bilayers \cite{akutsu81,altenbach84}.
However, the changes of the headgroup order parameters as a function of CaCl$_2$ 
concentration were overestimated by the same combination of force field parameters.

@@

%To improve the cation binding behaviour to POPC lipid bilayer we implicitly included the electronic
%polarization to the Lipid14 model by using the ECC approach as described in section \ref{section:ecc}.
Our simulations with the ECC-lipid and the ECC-ion models \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}
exhibit an improved behaviour of cation binding to a POPC bilayer
showing a good agreement with experiments in the changes of 
the lipid headgroup order parameters as a function of NaCl and CaCl$_2$ concentrations
in Fig.~\ref{fig:delta_ordPar_CaCl}. Since also the headgroup order parameter
response the to the bound positive charge the ECC-lipid model
was in good agreement with experiments in section \ref{section:boundCHARGE},
we conclude that the model correctly reproduces the binding affinity of
Na$^{+}$ and Ca$^{2+}$ ions to POPC lipid bilayer.
%In conclusion, the results suggest that 
%the ECC-lipid model together with the ECC-ion parameters gives a realistic
%description for the binding affinities of Na$^+$ and Ca$^{2+}$ ions to POPC lipid bilayer. 
Furthermore, the overestimated headgroup order parameter changes of POPC in the Lipid14
model as a function of CaCl$_2$ concentration arise from both, the overestimated
binding affinity and sensitivity of the headgroup tilt to the bound positive charge.
This probably applies also to the other lipid models considered in the previous study~\cite{catte16},
emphasizing the importance of the comparison of the lipid headgroup order parameter response
to the bound charge between simulations and experiments as done in section \ref{section:boundCHARGE}.

\subsection{Cation binding to POPC membrane in atomistic detail}

To quantify the binding affinities in different models we calculated 
the relative surface excess from Eq.~\ref{surfexcess}. 
Calcium cations at a molar concentration of $350\,\mathrm{mM}$ provide
$\Gamma_i^w = 0.07 \pm 0.01 \rm{nm}^{-2}$ in the ECC-lipid model, which is significantly
smaller than the values for the Lipid14 model with \AA{}qvist ions, $\Gamma_i^w = 0.13 \pm 0.01 \rm{nm}^{-2}$,
or with Dang ions, $\Gamma_i^w = 0.3 \pm 0.03 \rm{nm}^{-2}$.
\todo{JOE: I'd put the $\Gamma$ values into a table}
Interestingly, the relative surface excess of 
\ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) is qualitatively different from \ce{CaCl2}
having a negative value $\Gamma_i^w = -0.1 \pm 0.01 \rm{nm}^{-2}$, 
meaning that water molecules are preferred to sodium and chloride ions at the membrane-water interface. 
This is in contrast to previous works employing classical MD simulations,
which generally agree together on the existence of an increased sodium density at the membrane-water interface.
\todo{JOE: Add references to NaCl simulation papers that show peaks.
Add a simple note that there is NBFIX trick in Charmm36 (?).} 
This finding is depicted in Figures~\ref{fig:nacl-dens} and~\ref{fig:cacl-dens},
which show the density profile of ions and water along the membrane normal. 
%The density profiles show
%a larger Ca$^{2+}$ density peak in lipid headgroup region for
%the Lipid14 model with Dang and ECC-ions than for the ECC-lipid model.
\todo{Densities and relative surface excess from NaCl systems to be added.
  The discussion below to be finished after this.}

@@

We can estimate relative binding affinities 
of several moieties in POPC towards \ce{Ca^{2+}} and \ce{Na+} cations 
using the number of contacts per lipid between cations and POPC oxygen atoms, 
$\Sigma ^\mathrm{Na^{2+}} _\mathrm{O}$,
in which a contact between the two atoms is counted 
if their distance is less than or equal $0.3 \, \mathrm{nm}$ 
(robustness of this definition discussed in reference~\citenum{javanainen17}).

For sodium at $1000\,mM$ concentration 
the average number of contacts per lipid with any oxygen atom in POPC is 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O} = 0.106 $,    % = 17.1/128 
whereas if only phosphate oxygen atoms are considered 
this number decreases to 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{PO_4} = 0.085 $,    % = 17.0/128 . 
and if only carbonyl oxygen atoms are counted 
it is even lower, 
$\Sigma ^\mathrm{Na^{+}} _\mathrm{O_{carb.}} = 0.048 $.    % = 4.3/128 . 
This means that interaction of sodium and POPC 
has balanced contributions of 55\% from only phosphate oxygens 
(i.e. no contacts with carbonyl oxyens),
and from only carbonyl oxygens with 20\% of configurations 
leaving 25\% for mixed phosphate and carbonyl oxygen configurations.  
This is in line with the calculated negative value of 
relative surface excess of sodium with respect to water, $\Gamma_i^w$, and
their density profiles along the membrane normal (Fig.~\ref{fig:nacl-dens})
suggesting that the populations of bound sodium cations 
naturally decrease with the decreasing density of water.

In contrast to sodium, the dominant contribution
to the binding of \ce{Ca^{2+}} to POPC membranes comes from the phosphate group. 
% Numbers from simulation with 346mM conc. CaCl2
The average number of contacts per lipid 
between \ce{Ca^{2+}} at $290\,\mathrm{mM}$ concentration and any oxygen atom in POPC is 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.134 $,    % = 17.1/128 
whereas if only phosphate oxygen atoms are considered 
the average number of contacts decreases only by a tiny amount to 
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.133 $,    % = 17.0/128 . 
and if only carbonyl oxygen atoms are considered 
%(phosphate oxygens contribute, but are not counted as contacts),
this value is merely
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.033 $.    % = 4.3/128 . 
% Numbers from simulation with 980/845mM conc. CaCl2
%The average number of contacts per lipid between \ce{Ca^{2+}} and any oxygen atom in POPC is 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $,    % = 33.8/128 
%whereas if only phosphate oxygen atoms are considered 
%the average number of contacts decreases only by a tiny amount to 
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{PO_4} = 0.262 $,    % = 33.5/128 . 
%and if only carbonyl oxygen atoms are considered 
%(phosphate oxygens contribute, but are not counted as contacts),
%this value is merely
%$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} = 0.066 $.    % = 8.4/128 . 
Hence, calcium is in contact with phosphate oxygens for 99\% of the time it is bound to the membrane, 
whereas only 25\% of the time accounts for contacts with carbonyl oxygens,
and interactions purely with carbonyl oxygens account for less than 1\% of configurations. 
Higher concentration of \ce{CaCl2} naturally increases the amount of contacts per lipid,
for $850\,\mathrm{mM}$ concentration of \ce{CaCl2}
the average number of contacts per lipid is increased to
$\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O} = 0.264 $, 
but the distribution of states between phosphate and carbonyl oxygens remains unchanged. 
We corroborate this finding with the probability isodensity contours in Fig.~\ref{fig:volmaps}, 
and density profiles of ions and water along the membrane normal in Fig.~\ref{fig:cacl-dens}.

\subsection{Residence times of \ce{Na+} and \ce{Ca^{2+}} cations in POPC membrane}

The residence time of \ce{Ca^{2+}} in a POPC membrane 
is experimentally estimated to be lower than $10\,\mu\mathrm{s}$~\cite{altenbach84}. 
From a recent theoretical work with long simulations \cite{javanainen17}
this time can be roughly estimated 
in the order of $1\,\mu\mathrm{s}$~\cite{javanainen17}. 
Based on Fig. S17 in that work, Ref.~\citenum{javanainen17},
we took the simulation with the apparent highest fluctuations of the number of bound calcium,
which can be obtained at \url{https://doi.org/10.5281/zenodo.259376}
and anaylsed the residence times.
% simulation: (Matti C36 with CaClMK, 800ns total):
We find out that
approximately 20\% of bound residence times of \ce{Ca^{2+}} 
are limited by the length of the simulation, $800\,\mathrm{ns}$, 
and only less than
60\% of bound residence time is smaller than half of the simulation length , $400\,\mathrm{ns}$. 
Other simulations reported in that work show even longer residence times.

This is in contrast to our model, ECC-lipids, 
which estimates the residence time of calcium bound to a POPC membrane to be 
smaller than $60\,\mathrm{ns}$ for 90\% of the time a cation is bound. % exactly $53\,\mathrm{ns}$
i.e. at least two orders of magnitude lower than previous estimates.
In addition, the longest observed residence time in our simulation is $120\,\mathrm{ns}$, 
which is well below the length of our simulations.
This is unfortunately not the case of many theoretical works,
where the residence time cannot be estimated,
or the simulation has not even decorrelated from it's 
initial condition with a population of bound/unbound cations \cite{catte16}, 
because it is limited by the simulation length, 
even though it can be very long (order of $\mu\mathrm{s}$) in some works, e.g. reference~\citenum{javanainen17}. 
Increasing time scales to the order of $\mu\mathrm{s}$ 
doesn't help to alleviate such a problem in current lipid models \cite{javanainen17}. %and also \cite{catte16}, but same simulation was used. 
Such a finding changes the point of view on binding of calcium to PC membranes from
a very strong long-term stable binding with rare exchanges to 
a relatively frequent exchange of clacium cations 
in equilibrium between membrane and solvent.

Interestingly, \ce{Na+} cations exhibit much more rapid exchange between membrane and solvent than calcium.
The residence time of sodium bound to a POPC membrane is
smaller than $2\,\mathrm{ns}$ for 90\% of the time it is bound to the membrane 
with the longest measured bound time being $13\,\mathrm{ns}$. 
Such an exchange between membrane and solvent 
is an order of magnitude faster than for calcium cations. 
A histogram of residence times from our simulations is in Fig.~S\ref{fig:hist_residence_times} in SI. 
\todo{JOE: Plot a histogram of residence times based on our simulations and simulation from \citenum{javanainen17} and put it into SI.
 }

\subsubsection{Internal side-note: }
However, $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$ is increased 
in simulation withy the highest \ce{CaCl2} concentration
when carbonyl dipoles are scaled only with 0.945 
(roughly done through charge redistribution)
 to : $\Sigma ^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} \approx 0.08 $.    % = (70.-59.8)/128 .

This means that calcium is in contact with a phosphate oxygen for 98\% of the time it is bound to the membrane, 
whereas 46\% of the time accounts for contact with a carbonyl oxygen,
and purely carbonyl bound states account for less than 2\% of configurations. 
The states that have a contribution from carbonyl oxygens also contain a bound phosphate, 
so it can be thought of rather as a stabilizing moiety.

The residence times change to 75ns (90\% of bound time) and max is 185ns.
That's not much, too. It certainly is not a game changer, I'd even call it to be within accuracy. 
The slight problem may be that with this tweak, there's approx. 17\%  more cations bound to the membrane,
which means, I'd need to tune the sigma parameter a bit (increase the scaling factor $f_\sigma$). 
That could detune the membrane structure \dots
Not $f_q$ -- that's set by the simulations with the surfactant. Good.
I think that such a binding detail is within the accuracy of the proposed simple correction.
It would be worth noting that the model may underestimate slightly the contribution from carbonyls due to this. 
But we can't say to what extent\dots

\subsection{Stoichiometry of binding of \ce{Na+} and \ce{Ca^{2+}} cations to POPC membrane}

@@

Simulations allow us to directly evaluate the stoichiometry 
by calculating relative propensities of various \ce{Ca^{2+}}:$n\times$POPC complexes
by evaluating contacts between cations and lipids with a cut off radius $0.3\,\mathrm{nm}$. 
The propensities of \ce{Ca^{2+}}:$n\times$POPC complexes at 285~mM concetration 
are in order:
$42\%$ for two lipids,
$30\%$ for one lipid, and
$28\%$ for three lipids.
Apart from a ternary complex, in which one calcium cation binds two lipids, 
we also find complexes with 1~and 3~lipids
occuring with only a slightly lower but similar probability
as shown in Figure~\ref{fig:cacl_complexes}.

@@

Several binding models were proposed and tested~\cite{altenbach84} of which only one,
ternary complex binding model, provided a good fit of the experimental observations. 
In such a model, it is assumed that \ce{Ca^{2+}} cations bind to a POPC membrane 
with a stoichiometry 2~POPC:1~\ce{Ca^{2+}}.
In a later work~\cite{macdonald87}, 
a Langmuir adsorption model (i.e. stoichiometry 1~POPC:1~\ce{Ca^{2+}}) 
was found to provide as good fit as ternary complex model, 
when only low concentrations of \ce{CaCl2} are considered.
Ternary complex model also provides a good fit 
to our simulations with ECC-lipids 
(see Fig.~\ref{fig:cacl-bind} in SI and its caption for details). 
The symmetry of the distribution of complexes from simulation 
as shown in Figure~\ref{fig:cacl_complexes} --   
i.e. almost equal probabilities of complexes with 1 or 3 lipids 
that behave in the total average picture as complexes with 2 lipids -- 
provides clues why ternary complex binding model 
fits both simulation and experimental results relatively well, 
although it is apparently incorrect. 
It is also worth noting that many binding constants in the literature
(e.g., Marsh's handbook \cite{marsh13}) 
are based on the headgroup order parameter changes 
intepreted with the ternary complex binding model.

\todo{SAMULI: I think the we should make clear that the current binding
  constants in, e.g., Marsh's handbook are based on the headgroup order parameter
  changes intepreted with ternary complex binding model. I.e. the experimental raw data
  is excatly the same as we have in this work. Our simulations enable much more versatile interpretation
  of the binding phenomena. I think that this is one the main points of this work, so this comment
  applies to the previous section, introduction, abstract and conclusions as well.
  JOE: agree, this would be a good statement in general. I'll try to put it into the text.
}

Binding of \ce{Na+} cations to a POPC membrane 
mostly follows stoichiomtetry \ce{Na^{+}}:$1\times$POPC 
as shown in Figure~\ref{fig:cacl_complexes}.

\section{Conclusions}

By using the electrometer concept we show that the binding of \ce{Na^+} and \ce{Ca^2+} ions
to a POPC lipid bilayer can be accurately described with the classical MD simulation 
force field, where the electronic polarization is implicitly included using 
the electronic continuum correction (ECC)~\cite{leontyev11}.
The proposed ECC-lipids model is a significant improvement over 
other available lipid models, which all overestimate specific cation binding affinities~\cite{catte16}.  
While the structural details of a POPC lipid bilayer simulated with the ECC-lipids
model agree with experiments with the comparable accuracy to the other state of the art lipid models,
it also reproduces the experimental lipid head group order parameter responses to
the cationic surfactant, NaCl and CaCl$_2$ concentrations.
